\bf On Misuses of the Kolmogorov--Smirnov Test for One-Sample Goodness-of-Fit

### 2024-03-30 manuscript/manuscript.tex
added classroom implementation at graduate level

% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}

@@

\begin{abstract}
The Kolmogorov--Smirnov (KS) test is widely employed to assess the
goodness-of-fit of a hypothezized continuous distribution to a sample.
Despite its popularity, the test is frequently misused in the literature and
practice. While originally intended for independent, continuous data with
precisely specified hypothesized distributions, it is erroneously applied to
scenarios with dependent, discrete, or rounded data, with hypothesized
distributions requiring estimated parameters. For example, it has been
``discovered'' multiple times that the test is too conservative when the 
hypothesized distribution has parameters that need to be estimated.
We demonstrate misuses of the one-sample KS test in
three scenarios through simulation studies:
1) the hypothesized distribution has unspecified parameters;
2) the data are serially dependent; and
3) a combination of the first two scenarios.
For each scenario, we provide remedies for practical applications using
appropriate bootstrap approaches.
The whole demonstration can be used as hands-on education materials on both
goodness-of-fit tests and bootstrap.

@@

The standard KS test does not apply to stationary yet dependent data either. The
distribution of the KS statistic would have a higher variance for positively
dependent data than that derived when the data are independent because of a
smaller effective sample size. For example, for testing normality,
\citet{durilleul1992lack} demonstrates that a naive application of
the KS statistic is too liberal for medium-to-high positive serial dependence,
and that for negative dependence, the behavior is asymmetrical. For remedies,
\citet{weiss1978modification} provides a procedure that is applicable 
specifically for data
modeled by the second-order auto-regressive (AR) process where the AR parameters
are known. Serial dependence affects the validity of the two-sample KS test too.
In a recent paper on online diagnosis of performance variation in
high-performance computing systems, for example, the authors provided no details
about whether they accounted for serial dependence when applying the two-sample
KS test \citep{tuncer2019online}. After comparing various strategies for dealing
with serial dependence, \citet{lanzante2021testing} concludes that a test based
on Monte-Carlo simulations performed the best.
When additionally the hypothesized distribution contains unknown
parameters, the standard KS test becomes even further inapplicable.

@@

The consequence of using the wrong null distribution for the KS statistic $D_n$
in~\eqref{eq:ks_fitted} can be illustrated through a simple simulation study. A
random sample $X_1, \ldots, X_n$ was generated from a normal
distribution with both mean and variance parameters set equal to 8 ($N(8,8)$), 
and sample size $n = 200$. The maximum likelihood estimates of the mean and 
variance were used as $\hat\theta_n$. The p-value of the test
statistic~$D_n$ in Equation~\eqref{eq:ks_fitted} was obtained by naively calling 
the \texttt{ks.test()} function in~R with the fitted normal distribution as the
hypothesized distribution. That is, the hypothesized distribution was
$N(\bar X, s^2)$ where $\bar X$ is the sample mean and $s^2$ is the sample
variance (with $n$ in the denominator). This experiment was repeated 1000 times
and the probability-probability plot (PP-plot) of the 1000 p-values are 
displayed in the top left plot of Figure~\ref{fig:pp_f}. If the test were
valid, the p-values would be uniformly distributed in the unit
interval and the points in the PP-plot would fall along the diagonal line. 
Clearly, the distribution of the 1000 p-values are very different from what
one would expect to see from 1000 draws from the standard uniform
distribution.

@@

We similarly considered random samples $X_1, \ldots, X_n$ 
generated from a gamma distribution with shape and scale parameters 8 and 1, 
respectively ($\Gamma(8,1)$), and sample size $n=200$. These parameter values 
were selected so that mean and variance of the normal and gamma distributions 
in the simulations are the same. Likewise, the p-value of the test 
statistic~$D_n$ in~\eqref{eq:ks_fitted} was obtained by naively calling the
\texttt{ks.test()} function in~R with the fitted gamma distribution as the
hypothesized distribution using the shape and scale maximum likelihood 
estimates. The PP-plot of the p-values from 1000 replications of this 
experiment are displayed in the bottom left plot of Figure~\ref{fig:pp_f}. The
p-values are again non-uniformly distributed due to the use of the 
estimated parameters when specifying the null distribution.

@@

Here we consider the situation where the observed data $X_1, \ldots, X_n$ are
not independent but serially dependent, and the hypothesized distribution $F$
contains no unknown parameters. That is,
\[
H_0: \text{$X_i$'s have marginal distribution $F$.}
\]
The KS statistic $D_n$ in Equation~\eqref{eq:ks_standard}
is still a good measure of deviation from the null hypothesis. Nonetheless, the
null distribution of $F_n$ changes when there is serial dependence.
When the serial dependence is
positive, the effective sample size is less than~$n$. The test statistic
would be stochastically greater than that obtained from independent data, 
resulting in a test that is too liberal (i.e., the null hypothesis is rejected
more often than it should). The reasoning is similar when the serial dependence 
is negative, in which case, the test is conservative.

@@

@@

The three kinds of misuses of the KS test for one-sample goodness-of-fit that we
demonstrated here are commonly seen in practice. With the aid of computing tools
such as R, they can be easily demonstrated in classroom teaching. Practitioners
need to be aware of the assumptions of the standard one-sample KS test that the
hypothesized distribution is completely specified and that the data are
independent. When these assumptions are not met, naive applications of the test
are no longer accurate and remedies must be performed. In the
case of fitted parameters, parametric and nonparametric bootstrap can restore
the size of the test. The nonparametric version has long been developed too but
is lesser known \citep{babu2004goodness}. In the case of serially dependent data,
our semiparametric bootstrap procedure with a working dependence structure
controlled by an automatically selected ARMA model restores the size of the test
for a wide range of dependence structures.
When both assumptions are violated, i.e., where the data has serial 
dependence and the parameters must be fitted, a combination of the two
individual remedies shows good results. The misuses of the KS tset also apply
to other goodness-of-fit tests based on the empirical distribution
function, such as the Anderson--Darling test or the Cramér-—von Mise test, which
are known to be superior to KS in certain respects \citep{stephens2017tests}.
Similar bootstrap remedies can be developed for these tests.

@@

@@

\section{Nonparametric Bootstrap for KS Test with Fitted Parameters}

>>>>

% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

@@

\begin{abstract}
The Kolmogorov--Smirnov (KS) test is widely employed to assess the
goodness-of-fit of a hypothesized continuous distribution to a sample.
Despite its popularity, the test is frequently misused in the literature and
practice. While originally intended for independent, continuous data with
precisely specified hypothesized distributions, it is erroneously applied to
scenarios with dependent, discrete, or rounded data, with hypothesized
distributions requiring estimated parameters. For example, it has been
``discovered'' multiple times that the test is too conservative when the 
hypothesized distribution has parameters that need to be estimated.
We demonstrate misuses of the one-sample KS test in
three scenarios through simulation studies:
1) the hypothesized distribution has unspecified parameters;
2) the data are serially dependent; and
3) a combination of the first two scenarios.
For each scenario, we provide remedies for practical applications using
appropriate bootstrap approaches.
The whole demonstration can be used as hands-on education materials on both
goodness-of-fit tests and bootstrap.

@@

The standard KS test does not apply to stationary yet dependent data either. The
distribution of the KS statistic would have a higher variance for positively
dependent data than that derived when the data are independent because of a
smaller effective sample size. For example, for testing normality,
\citet{durilleul1992lack} demonstrate that a naive application of
the KS statistic is too liberal for medium-to-high positive serial dependence,
and that for negative dependence, the behavior is asymmetrical. For remedies,
\citet{weiss1978modification} provides a procedure that is applicable 
specifically for data
modeled by the second-order auto-regressive (AR) process where the AR parameters
are known. Serial dependence affects the validity of the two-sample KS test too.
In a recent paper on online diagnosis of performance variation in
high-performance computing systems, for example, the authors provided no details
about whether they accounted for serial dependence when applying the two-sample
KS test \citep{tuncer2019online}. After comparing various strategies for dealing
with serial dependence, \citet{lanzante2021testing} concludes that a test based
on Monte-Carlo simulations performed the best.
When additionally the hypothesized distribution contains unknown
parameters, the standard KS test becomes even further inapplicable.

@@

The consequence of using the wrong null distribution for the KS statistic $D_n$
in~\eqref{eq:ks_fitted} can be illustrated through a simple simulation study. A
random sample $X_1, \ldots, X_n$ was generated from a normal
distribution with both mean and variance parameters set equal to 8 ($N(8,8)$), 
and sample size $n = 200$. The maximum likelihood estimates of the mean and 
variance were used as $\hat\theta_n$. The p-value of the test
statistic~$D_n$ in Equation~\eqref{eq:ks_fitted} was obtained by naively calling 
the \texttt{ks.test()} function in~R with the fitted normal distribution as the
hypothesized distribution. That is, the hypothesized distribution was
$N(\bar X, s^2)$ where $\bar X$ is the sample mean and $s^2$ is the sample
variance (with $n$ in the denominator). This experiment was repeated 1000 times
and the probability-probability plot (PP-plot) of the 1000 p-values are 
displayed in the top left plot of Figure~\ref{fig:pp_f}. If the test were
valid, the p-values would be uniformly distributed in the unit
interval and the points in the PP-plot would fall along the diagonal line. 
Clearly, the distribution of the 1000 p-values is very different from what
one would expect to see from 1000 draws from the standard uniform
distribution.

@@

We similarly considered random samples $X_1, \ldots, X_n$ 
generated from a gamma distribution with shape and scale parameters 8 and 1, 
respectively ($\Gamma(8,1)$), and sample size $n=200$. These parameter values 
were selected so that the mean and variance of the normal and gamma distributions 
in the simulations are the same. Likewise, the p-value of the test 
statistic~$D_n$ in~\eqref{eq:ks_fitted} was obtained by naively calling the
\texttt{ks.test()} function in~R with the fitted gamma distribution as the
hypothesized distribution using the shape and scale maximum likelihood 
estimates. The PP-plot of the p-values from 1000 replications of this 
experiment are displayed in the bottom left plot of Figure~\ref{fig:pp_f}. The
p-values are again non-uniformly distributed due to the use of the 
estimated parameters when specifying the null distribution.

@@

Here we consider the situation where the observed data $X_1, \ldots, X_n$ are
not independent but serially dependent, and the hypothesized distribution $F$
contains no unknown parameters. That is,
\[
H_0: \text{$X_i$'s have marginal distribution $F$.}
\]
The KS statistic $D_n$ in Equation~\eqref{eq:ks_standard}
is still a good measure of deviation from the null hypothesis. Nonetheless, the
null distribution changes when there is serial dependence.
When the serial dependence is
positive, the effective sample size is less than~$n$. The test statistic
would be stochastically greater than that obtained from independent data, 
resulting in a test that is too liberal (i.e., the null hypothesis is rejected
more often than it should). The reasoning is similar when the serial dependence 
is negative, in which case, the test is conservative.

@@

\section{Classroom Implementation}

% Since this is a Teacher's Corner submission, I was hoping to read about how
% these ideas have been presented in specific courses and also how students
% reacted to this. What background did the students have? What level of detail was
% given to them? That is, it is one thing to say "use the bootstrap" and another
% thing to present to students, e.g., a Markdown file with all necessary code
% included. Was that done? Were students invited to conduct their own simulations?
% Etc.

% Graduate level
The misuses of one-sample KS test can be easily incorporated into graduate level
classroom teaching for courses that cover parametric bootstrap. For example, at
the University of Connecticut (UConn), JY has always included a demonstration of
the misuse under unspecified parameters in STAT 5361, Statistical Computing. The
catalog description of this course includes is: ``Use of computing for statistical
problems; obtaining features of distributions, fitting models and implementing
inference. Basic numerical methods, nonlinear statistical methods, numerical
integration, modern simulation methods.'' Bootstrap is covered under ``modern
simulation methods.''  The course is open to graduate students in Statistics,
others with permission. Detailed R code and graphical illustrations available in
the open-access course notes elucidate the distinctions between employing null
hypotheses with and without defined parameters (accessible at:
\url{https://jun-yan.github.io/stat-5361/boot.html#parametric-bootstrap}).
Students were able to follow the demonstration closely, gaining a clear
understanding of the implications of using estimated parameters within test
statistics, and learning about the efficacy of the parametric bootstrap method
as a corrective measure.

Supplementary materials accompanying this paper present R code that simplifies
the illustration of the three common misuses of the one-sample KS test in
classrooms, whether during lectures or as part of homework assignments.
Addressing the misuses in contexts of serial dependency necessitates familiarity
with ARMA models in time series analysis---a prerequisite achievable by most
statistics graduate students. While incorporating copulas to model dependency
structures extends beyond conventional statistics curricula, it presents an
opportunity for advanced exploration in courses focused on statistical computing
or multivariate analysis.

% Undergraduate level (could mention AZ's undergraduate thesis)

@@

The three kinds of misuses of the KS test for one-sample goodness-of-fit that we
demonstrated here are commonly seen in practice. With the aid of computing tools
such as R, they can be easily demonstrated in classroom teaching. Practitioners
need to be aware of the assumptions of the standard one-sample KS test that the
hypothesized distribution is completely specified and that the data are
independent. When these assumptions are not met, naive applications of the test
are no longer accurate and remedies must be performed. In the
case of fitted parameters, parametric and nonparametric bootstrap can restore
the size of the test. The nonparametric version has long been developed too but
is lesser known \citep{babu2004goodness}. In the case of serially dependent data,
our semiparametric bootstrap procedure with a working dependence structure
controlled by an automatically selected ARMA model restores the size of the test
for a wide range of dependence structures.
When both assumptions are violated, i.e., where the data has serial 
dependence and the parameters must be fitted, a combination of the two
individual remedies shows good results. The misuses of the KS test also apply
to other goodness-of-fit tests based on the empirical distribution
function, such as the Anderson--Darling test or the Cramér-—von Mise test, which
are known to be superior to KS in certain respects \citep{stephens2017tests}.
Similar bootstrap remedies can be developed for these tests.

@@

\section*{Data Availability Statement}
An R implementation of the proposed methods will be made publicly available upon 
publication.

@@

\section{Appendix: Nonparametric Bootstrap for KS Test with Fitted Parameters}
